\section{Bodart's Virtually Engel Witness}
\label{sec:VirtuallyEngel}

The witness is Bodart's virtually Engel group. It is presented by two generators $a,t$ and four
relators. Writing
\[
a^t=t^{-1}at
\qquad\text{and}\qquad
C=[a,[a,a^t]],
\]
the defining relations are
\[
t^2=1,\qquad
C=[a^t,[a,a^t]],\qquad
[C,a]=1,\qquad
[C,a^t]=1.
\]
The finite alphabet used for geodesic growth is
\[
S=\{a,a^{-1},t\},
\]
interpreted in the presented group in the evident way.

\subsection*{Internal coordinate model}
The current development also fixes an explicit computable target for this presentation. One
introduces integer coordinates
\[
(u,v,A,B)\in \mathbf Z^4,
\]
where $u,v$ record the two diagonal directions in the Engel subgroup, $A$ records the algebraic
area term, and $B$ records three times the barycentric term used in Bodart's path model. The
multiplication law is
\[
(u,v,A,B)(u',v',A',B')
=
\bigl(
u+u',\,
v+v',\,
A+A'+vu'-uv',\,
B+B'+3(u-v)A' +(2(u-v)+(u'-v'))(vu'-uv')
\bigr).
\]
This yields a concrete nilpotent group of Engel type. Reflection in Bodart's picture induces the
order-two automorphism
\[
\sigma(u,v,A,B)=(-v,-u,-A,-B).
\]
The resulting semidirect product with the order-two symmetry is the computable coordinate model
used in the current Lean development.

The generators $a$ and $t$ of the presentation are sent respectively to the diagonal step
\[
(1,0,0,0)
\]
inside the nilpotent factor and to the non-trivial symmetry. The four defining relators are then
verified directly in this coordinate model, so the presented group admits a canonical homomorphism
into this explicit semidirect product.

The subgroup generated by the diagonal steps is smaller than the ambient coordinate group. In the
current Lean development this subgroup is cut out internally by two congruence conditions on the
coordinates:
\[
A+uv\equiv 0 \pmod 2,
\]
and
\[
B-3A(u-v)-u^{2}v+2uv^{2}\equiv 0 \pmod 6.
\]
These conditions are proved to be stable under multiplication, inversion, and the reflection
automorphism. This isolates the lattice generated by the diagonal directions inside the nilpotent
coordinate model and is the present starting point for the internal word-problem argument.

The current development also corrects the commutator used in the normal-form coordinates. Since
mathlib writes
\[
[x,y]=xyx^{-1}y^{-1},
\]
the literal word $[a,a^t]$ lands at a different barycentric coordinate from the Hall-style
commutator
\[
a^{-1}(a^t)^{-1}a a^t.
\]
It is this Hall commutator that matches the experimentally derived normal forms. Using it, the
Lean development now assigns to every lattice point an explicit computable Hall normal-form word
and proves that its image in the coordinate model recovers the original lattice point exactly.

\subsection*{Imported literature facts}
The first imported input is Bodart's asymptotic theorem:
\[
\gamma_{\mathrm{geod}}(n) \asymp \exp(n^{3/5}\log n).
\]
This gives intermediate geodesic growth and, in particular, excludes rational geodesic growth for
the chosen alphabet once the abstract asymptotic-to-irrationality bridge has been formalized.

The second imported input is the standard consequence that this virtually nilpotent group has
solvable word problem. Thus one obtains the following two facts:
\begin{enumerate}[label=(\roman*)]
  \item the virtually Engel presentation has solvable word problem;
  \item the geodesic growth with respect to $S=\{a,a^{-1},t\}$ is irrational.
\end{enumerate}
At present these two witness properties are still imported. The new coordinate model is the
internal starting point for the word-problem part of the argument.
